We are not the 99 percent: quantifying asphericity in the distribution of Local Group satellites

### 2018-01-24 paper/paper.tex
edit

\begin{abstract}
We quantify the asphericity in the spatial distribution of the
brightest satellites around the Milky Way (MW) and M31, the dominant
galaxies in the Local Group (LG).
We use simulations to build an explicit probability distribution for the
asphericity around galaxies similar to the LG in the Lambda Cold Dark
Matter (LCDM) paradigm.
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-DarkMatter and ELVIS. 
Detailed results differ greatly among the simulations, suggesting a
strong influence of resolution, modelled physics and criteria to select
LG pairs.
However, there are three common trends in these simulations. First, at most of
$1\%$ the pairs are expected to have satellite distributions with the
same asphericity as the Local Group; second, between
$38\%$ to $59\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $2\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in M31 galaxy turn out to have a rather typical distribution 
in the LCDM context, while the MW is atypical in this respect. 
These results place the LG close to the level of a 3-$\sigma$ outlier
in the LCDM paradigm.
We summarize our results in terms of mean values and covariance
matrices. 
This approach facilitates the comparison between asphericity results
from different numerical setups and choices to select LG pairs in
simulations.  
\end{abstract}

@@

We extract halo and galaxy information from the Illustris-1 simulation
which has the highest resolution in the current release of the
Illustris Project.
Illustris-1 has $1820^3$ dark matter particles and $1820^3$ initial gas
volumen elements. 
This corresponds to a dark matter particle mass of
$6.3\times 10^6$\Msun\ and a minimum mass for the baryonic volume
element of $8.0\times 10^7$\Msun. 
The corresponding spatial resolution is $1.4$ kpc for the dark matter
gravitational softening and $0.7$ kpc for the typical size of the
smallest gas cell size.

@@

\begin{table*}
  \centering
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{|p{2.5cm}|c|c|c|c|c|c|}
    \hline
    \multirow{2}{4.0cm}{} & \multicolumn{2}{c|}{\textbf{Illustris-1-DM}} & \multicolumn{2}{c|}{\textbf{Illustris-1}} & \multicolumn{2}{c|}{\textbf{ELVIS}}\\
    % \hline
    % \textbf{Inactive Modes} & \textbf{Description}\\
    \cline{2-7}
    & \textbf{M31} & \textbf{MW} & \textbf{M31} & \textbf{MW}& \textbf{M31} & \textbf{MW}\\
    %\hhline{~--}
    \hline
    Plane width (kpc) & $62\pm 3$   & $61\pm 3$     & $70\pm 4$ & $67\pm 2$ & $70\pm 2$& $68\pm 4$ \\\hline
    $c/a$ ratio & $0.50\pm0.01$ & $0.50\pm 0.02$ & $0.52\pm 0.01$ & $0.53\pm 0.01$ & $0.54\pm 0.01$& $0.49\pm 0.02$ \\ \hline
    $b/a$ ratio & $0.79\pm0.01$ & $0.81\pm 0.02$ & $0.80\pm 0.01$ & $0.80\pm 0.02$ & $0.80\pm0.01$& $0.81\pm 0.01$\\ \hline
  \end{tabular}
  \caption{Results from simulations. Mean values and standard deviations for the different
    quantities describing satellite distributions: plane width, $c/a$
    ratio and $b/a$ ratio. 
    The first column summarize the results from the 24 pairs in
    Illustris-1-DM, the second column from the 20 pairs in
    Illustris-1 and the third column from the 12 pairs in the
    ELVIS project.\label{table:simulations}}
\end{table*}

>>>>

\begin{abstract}
We quantify the asphericity in the spatial distribution of the
brightest satellites around the Milky Way (MW) and M31, the dominant
galaxies in the Local Group (LG).
We use simulations to build an explicit probability distribution for the
asphericity around galaxies similar to the LG in the Lambda Cold Dark
Matter (LCDM) paradigm.
This allows us to estimate the atipicallity
of the satellite distributions in the LG even when the underlying
simulations do not have enough systems that fully resemble the LG.
We demonstrate the method using three different simulations:
Illustris-1,  Illustris-1-Dark and ELVIS. 
Detailed results differ greatly among the simulations, suggesting a
strong influence of resolution, modelled physics and criteria to select
LG pairs.
However, there are three common trends in these simulations. First, at most of
$1\%$ the pairs are expected to have satellite distributions with the
same asphericity as the Local Group; second, between
$38\%$ to $59\%$ of the pairs have a halo with a satellite
distribution as aspherical as in M31; and third, at most $2\%$ of the
pairs have a satellite distribution as planar as in the MW. 
The brightest satellites in M31 galaxy turn out to have a rather typical distribution 
in the LCDM context, while the MW is atypical in this respect. 
These results place the LG close to the level of a 3-$\sigma$ outlier
in the LCDM paradigm.
We summarize our results in terms of mean values and covariance
matrices. 
This approach facilitates the comparison between asphericity results
from different numerical setups and choices to select LG pairs in
simulations.  
\end{abstract}

@@

We extract halo and galaxy information from the Illustris-1 and
Illustris-1-Dark simulations, the former simulation includes
hydrodynamics and star formation prescriptions while the latter only
includes dark matter physics.
These simulations have the highest resolution in the current release of the
Illustris Project.
Illustris-1 has $1820^3$ dark matter particles and $1820^3$ initial gas
volumen elements, while Illustris-1-Dark has $1820^3$ dark matter particles.
This corresponds to a dark matter particle mass of
$6.3\times 10^6$\Msun\ and a minimum mass for the baryonic volume
element of $8.0\times 10^7$\Msun for Illustris-1 and a dark matter
particle mass of $7.6\times 10^6$\Msun.
The corresponding spatial resolution is $1.4$ kpc for the dark matter
gravitational softening and $0.7$ kpc for the typical size of the
smallest gas cell size.

@@

\begin{table*}
  \centering
  \renewcommand{\arraystretch}{1.2}
  \begin{tabular}{|p{2.5cm}|c|c|c|c|c|c|}
    \hline
    \multirow{2}{4.0cm}{} & \multicolumn{2}{c|}{\textbf{Illustris-1-Dark}} & \multicolumn{2}{c|}{\textbf{Illustris-1}} & \multicolumn{2}{c|}{\textbf{ELVIS}}\\
    % \hline
    % \textbf{Inactive Modes} & \textbf{Description}\\
    \cline{2-7}
    & \textbf{M31} & \textbf{MW} & \textbf{M31} & \textbf{MW}& \textbf{M31} & \textbf{MW}\\
    %\hhline{~--}
    \hline
    Plane width (kpc) & $62\pm 3$   & $61\pm 3$     & $70\pm 4$ & $67\pm 2$ & $70\pm 2$& $68\pm 4$ \\\hline
    $c/a$ ratio & $0.50\pm0.01$ & $0.50\pm 0.02$ & $0.52\pm 0.01$ & $0.53\pm 0.01$ & $0.54\pm 0.01$& $0.49\pm 0.02$ \\ \hline
    $b/a$ ratio & $0.79\pm0.01$ & $0.81\pm 0.02$ & $0.80\pm 0.01$ & $0.80\pm 0.02$ & $0.80\pm0.01$& $0.81\pm 0.01$\\ \hline
  \end{tabular}
  \caption{Results from simulations. Mean values and standard deviations for the different
    quantities describing satellite distributions: plane width, $c/a$
    ratio and $b/a$ ratio. 
    The first column summarize the results from the 24 pairs in
    Illustris-1-Dark, the second column from the 20 pairs in
    Illustris-1 and the third column from the 12 pairs in the
    ELVIS project.\label{table:simulations}}
\end{table*}
